Do researchers preferentially collaborate with same-gendered colleagues?

### 2018-04-27 Gender homophilly project.Rmd
More writing

---
title: "Homophily in science: do researchers preferentially collaborate with same-gendered colleagues?"
subtitle: Online Supplementary Material
author: "Luke Holman and Claire Morandin"
output:
  html_document: 
    toc: true # show a table of contents
    toc_float: true # make table of contents float at the side of the page
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  # set to true if you want to number sections at each table header
    theme: lumen # Luke's favourite fonts and colours
    code_folding: hide # Add some buttons to show/hide the code
---

@@

```{r echo=FALSE, results='hide'}
library(knitr) # set up cacheing to save time when re-building the html document - chunks are not re-evaluated if the code has not changed
figure.counter <- 2  # Counters for numbering figures and tables
supp.figure.counter <- 1
table.counter <- 1
supp.table.counter <- 1
```
`r opts_chunk$set(cache=TRUE)`

@@

# This version uses bootstrapping to also estimate the 95% confidence limits on alpha
# Note that papers with lots of authors are more likely to be picked, ensuring that 
# each individual author is equally likely to end up in the sample. So if
# 2-author vs 10-author papers have different amounts of homophilly, or 
# different gender ratios, the CIs should not be biased
find.alpha.CIs <- function(nMales, nFemales, nBoots, correction = NULL){

@@

```{r}
minimum.n <- 50   # Only include journals with at least 50 papers in the specified range of time
unique.journals <- pubmed_sqlite %>% 
  select(journal) %>% distinct() %>% 
  collect(n = Inf) %>% .$journal # Get a list of all the journals in the database

@@

make.sample.size.table("older.papers") %>% 
  rename(`n (2005-2006)` = Value) %>% 
  left_join(make.sample.size.table("last.year") %>% 
              rename(`n (2015-2016)` = Value), by = "Quantity") %>% 
    pander(split.cell = 40, split.table = Inf)
```

@@

```{r}
# Calculate alpha across all authorship positions for all suitable NEW papers
gender.grouping.test.many.journals(readRDS("last.year.rds"),
                                   "all",
                                   "homophily.results.all.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

@@

```{r fig.width=5, fig.height=8}
figure2 <- master.dataset %>%
  filter(Age == "New") %>% 
  mutate(Significant = 1 * (adjusted.p.value < 0.05),
         Significant = replace(Significant, Significant == 0, "No"),
         Significant = replace(Significant, Significant == 1, "Yes")) 
n.sig.homo <- sum(figure2$Significant[figure2$Subset == "All authors"] == "Yes" & 
                    figure2$alpha.adjusted[figure2$Subset == "All authors"] > 0)
n.sig.hetero <- sum(figure2$Significant[figure2$Subset == "All authors"] == "Yes" & 
                    figure2$alpha.adjusted[figure2$Subset == "All authors"] < 0)
n.journals <- sum(figure2$Subset == "All authors")

figure2 <- figure2 %>%
  ggplot(aes(x = alpha.adjusted, fill = Significant)) + 
  geom_vline(xintercept = 0, linetype = 2) +
  geom_histogram(binwidth = 0.01, colour = "black") + 
  scale_fill_manual(values = c("skyblue", "white")) +
  xlab("Coefficient of homophily (\u03B1')") + 
  ylab("Number of journals") +
  facet_wrap(~Subset, scales = "free_y", ncol=1) + 
  theme_minimal(12) + 
  theme(legend.position = "top")
ggsave(plot = figure2, filename = "figures/figure2.pdf",  width = 5, height = 8, device = cairo_pdf)
figure2
```
<br></br> 
**Figure `r figure.counter`**: Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after adjusting p-values using Benjamini-Hochberg false discovery rate correction. The white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only.

```{r echo=FALSE, results='hide'}
figure.counter <- figure.counter + 1

# Save these figures for manuscript too:
saveRDS(c("n.journals" = n.journals, "n.sig.homo" = n.sig.homo, "n.sig.hetero" = n.sig.hetero), file ="manuscript/fig2_numbers.RData")
```

@@

gender.homophily.all %>% 
  filter(discipline %in% disc.summary$Discipline[disc.summary$`n journals` > 10]) %>%
  mutate(Discipline = factor(discipline, levels = rev(disc.summary$Discipline))) %>%
  ggplot(aes(x = Discipline, y = alpha.adjusted, fill = Discipline)) +
  geom_hline(yintercept = 0, linetype = 2) + 
  geom_boxplot(outlier.size = 0.5, outlier.shape = NA) + 
  geom_quasirandom(aes(colour = Discipline), alpha = 0.3, bandwidth = 0.25) + 
  ylab("Coefficient of homophily (\u03B1')") + 
  coord_flip() + 
  theme_minimal() + 
  theme(legend.position = "none")
```
<br></br>
**Figure `r figure.counter`**: Distribution of the $\alpha'$ values across journals, split by the research discipline of the journal. The points show the $\alpha'$ for each journal, and the boxes show quartiles 1, 2 and 3. The boxplot whiskers show the distance of the farthest point that is within 1.5 times the interquartile range from the edge of the box.

@@

### Gender homophily covaries with author gender ratio
Journals with a male-biased author gender ratio showed less author gender homophily, while journals with an even or female-biased author gender ratio showed more homophily. The pattern was similar whether we calculated $\alpha$ for all authors, first authors only, or last authors only. 
```{r fig.width=6, fig.height=5}
gender.ratio <- master.dataset %>% 
  filter(Age == "New") %>% 
  ggplot(aes(x = (gender.ratio.of.sample - 50), y = alpha.adjusted)) + 
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_point(alpha = 0.4, size = 0.7) + 
  stat_smooth(method = "gam", formula = y ~ s(x, k = 3), colour = "black", fill = "skyblue") + 
  xlab("Deviation from gender parity for all authors in this journal") + 
  ylab("Coefficient of homophily (\u03B1')") +
  facet_wrap(~Subset) + 
  theme_minimal() +
  theme(legend.position = "none") 
ggsave(plot = gender.ratio, filename = "figures/figure3.pdf", device = cairo_pdf, width=7, height=5)
gender.ratio
```
<br></br>
**Figure `r figure.counter`**: There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3.

@@

gender.homophily.all %>% 
  filter(!is.na(standardised.IF)) %>% 
  ggplot(aes(x = standardised.IF, y = alpha.adjusted)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  stat_smooth(method = "lm", colour = "black", fill = "skyblue") +
  geom_point(size = 0.5, alpha = 0.9) + 
  xlab("Standardised journal impact factor") + 
  ylab("Coefficient of homophily (\u03B1')") + 
  theme_minimal()
```
<br></br>
**Figure `r figure.counter`**: Journal impact factor (expressed relative to the average for the discipline) is negatively correlated with $\alpha'$. The relationship is weak ($R^2$ = `r R2`), but the results suggest that journals with strong homophily tend to have lower impact factors than journals with weak homophily in the same discipline.

@@

### Gender homophilly is also observed within countries

@@

```{r}
country.histo <- country.data %>%
  mutate(Significant = 1 * (p.adjusted < 0.05),
         Significant = replace(Significant, Significant == 0, "No"),
         Significant = replace(Significant, Significant == 1, "Yes")) %>%
  ggplot(aes(x = alpha.adjusted, fill = Significant)) + 
  geom_vline(xintercept = 0, linetype = 2) +
  geom_histogram(binwidth = 0.02, colour = "black") + 
  scale_fill_manual(values = c("skyblue", "white")) +
  xlab("Coefficient of homophily (\u03B1')") + 
  ylab("Number of journals") +
  theme_minimal() + 
  theme(legend.position = "top")
ggsave(plot = country.histo, filename = "figures/figure4.pdf", device = cairo_pdf, width = 6, height = 5)
country.histo
```
<br></br>
**Figure S`r supp.figure.counter`**: Histogram of $\alpha'$ for 325 unique combinations of journal and country, using data from August 2015 - August 2016. The white areas denote combinations for which $\alpha'$ differs significantly from zero (p < 0.05, following false discovery rate correction).

>>>>

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
subtitle: Online Supplementary Material
author: "Luke Holman and Claire Morandin"
output:
  html_document: 
    toc: true # show a table of contents
    toc_float: true # make table of contents float at the side of the page
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  # set to true if you want to number sections at each table header
    theme: lumen # Luke's favourite fonts and colours
    code_folding: hide # Add some buttons to show/hide the code
---

@@

```{r echo=FALSE, results='hide'}
library(knitr) # set up cacheing to save time when re-building the html document - chunks are not re-evaluated if the code has not changed
opts_chunk$set(warning=FALSE, message=FALSE) # Suppress warnings in the HTML output

figure.counter <- 2  # Counters for numbering figures and tables
supp.figure.counter <- 1
table.counter <- 1
supp.table.counter <- 1
```
`r opts_chunk$set(cache=TRUE)`

@@

# This version uses bootstrapping to also estimate the 95% confidence limits on alpha
# Note that papers with lots of authors are more likely to be picked, ensuring that 
# each individual author is equally likely to end up in the sample. So if
# 2-author vs 10-author papers have different amounts of homophily, or 
# different gender ratios, the CIs should not be biased
find.alpha.CIs <- function(nMales, nFemales, nBoots, correction = NULL){

@@

```{r get_data}
minimum.n <- 50   # Only include journals with at least 50 papers in the specified range of time
unique.journals <- pubmed_sqlite %>% 
  select(journal) %>% distinct() %>% 
  collect(n = Inf) %>% .$journal # Get a list of all the journals in the database

@@

sample.size.table <- make.sample.size.table("older.papers") %>% 
  rename(`n (2005-2006)` = Value) %>% 
  left_join(make.sample.size.table("last.year") %>% 
              rename(`n (2015-2016)` = Value), by = "Quantity")
write.csv(sample.size.table, file = "manuscript/sample.size.table.csv", row.names = FALSE)

sample.size.table %>% pander(split.cell = 40, split.table = Inf)
```

@@

```{r get_alpha}
# Calculate alpha across all authorship positions for all suitable NEW papers
gender.grouping.test.many.journals(readRDS("last.year.rds"),
                                   "all",
                                   "homophily.results.all.csv",
                                   nBoots = 1000, 
                                   nSimulations = 1000, 
                                   number.of.cores = 8,
                                   over.write = FALSE)

@@

```{r fig.width=5, fig.height=8}
figure2A <- master.dataset %>%
  filter(Age == "New") %>% 
  mutate(Significant = 1 * (adjusted.p.value < 0.05),
         Significant = replace(Significant, Significant == 0, "No"),
         Significant = replace(Significant, Significant == 1, "Yes")) 
n.sig.homo <- with(figure2A, sum(Significant[Subset == "All authors"] == "Yes" & 
                                   alpha.adjusted[Subset == "All authors"] > 0))
n.sig.hetero <- with(figure2A, sum(Significant[Subset == "All authors"] == "Yes" & 
                                     alpha.adjusted[Subset == "All authors"] < 0))
n.journals <- sum(figure2A$Subset == "All authors")

figure2A <- figure2A %>%
  ggplot(aes(x = alpha.adjusted, fill = Significant)) + 
  geom_vline(xintercept = 0, linetype = 2) +
  geom_area(binwidth = 0.01, colour = "black", stat = "bin") + 
  scale_fill_manual(values = c("skyblue", "white")) +
  xlab("Coefficient of homophily (\u03B1')") + 
  ylab("Number of journals") +
  facet_wrap(~Subset, scales = "free_y", ncol=1) + 
  theme_minimal(12) + 
  theme(legend.position = c(0.9, 0.96))
figure2A
```
<br></br>
**Figure 2A**: See the legend for Figure 2 below.

```{r echo=FALSE, results='hide'}
# Save these figures for manuscript too:
saveRDS(c("n.journals" = n.journals, "n.sig.homo" = n.sig.homo, "n.sig.hetero" = n.sig.hetero), file ="manuscript/fig2_numbers.RData")
```

@@

figure2B <- gender.homophily.all %>% 
  filter(discipline %in% disc.summary$Discipline[disc.summary$`n journals` > 10]) %>%
  mutate(Discipline = factor(discipline, levels = rev(disc.summary$Discipline))) %>%
  ggplot(aes(x = Discipline, y = alpha.adjusted)) +
  geom_hline(yintercept = 0, linetype = 2) + 
  geom_quasirandom(aes(colour = Discipline), alpha = 0.3, bandwidth = 0.25) + 
  geom_boxplot(outlier.size = 0.5, outlier.shape = NA, fill = NA) +
  ylab(NULL) + 
  coord_flip() + 
  theme_minimal() + 
  theme(legend.position = "none")

grid.arrange(figure2A + xlab(NULL), figure2B, nrow = 1, bottom = "Coefficient of homophily (\u03B1')")

ggsave(grid.arrange(figure2A + xlab(NULL), figure2B, nrow = 1, bottom = "Coefficient of homophily (\u03B1')"), filename = "figures/figure2.pdf", height = 8, width = 7, device = cairo_pdf)
```
<br></br>
**Figure `r figure.counter`**: Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after adjusting p-values using Benjamini-Hochberg false discovery rate correction. In the left panels, the white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (p < 0.05), while the blue area shows the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. The right panel shows the distribution of the $\alpha'$ values across journals, split by the research discipline of the journal. The points show the $\alpha'$ for individual journals, and the boxes show quartiles 1, 2 and 3. The boxplot whiskers show the distance of the farthest point that is within 1.5 times the interquartile range from the edge of the box.

@@

### Gender homophily covaries with author gender ratio
Journals with a male-biased author gender ratio showed less author gender homophily, while journals with an even or female-biased author gender ratio showed more homophily. The pattern was similar whether we calculated $\alpha$ for all authors, first authors only, or last authors only. 
```{r fig.width=6, fig.height=5}
gender.ratio <- master.dataset %>% 
  filter(Age == "New") %>% 
  ggplot(aes(x = (gender.ratio.of.sample - 50), y = alpha.adjusted)) + 
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_point(alpha = 0.4, size = 0.7) + 
  stat_smooth(method = "gam", formula = y ~ s(x, k = 3), colour = "tomato", fill = "skyblue") + 
  xlab("Deviation from gender parity for all authors in this journal") + 
  ylab("Coefficient of homophily (\u03B1')") +
  facet_wrap(~Subset) + 
  theme_minimal() +
  theme(legend.position = "none") 
ggsave(plot = gender.ratio, filename = "figures/figure3.pdf", device = cairo_pdf, width=7, height=5)
gender.ratio
```
<br></br>
**Figure `r figure.counter`**: There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3.

@@

figure4 <- gender.homophily.all %>% 
  filter(!is.na(standardised.IF)) %>% 
  ggplot(aes(x = standardised.IF, y = alpha.adjusted)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  stat_smooth(method = "lm", colour = "tomato", fill = "skyblue") +
  geom_point(size = 0.5, alpha = 0.9) + 
  xlab("Standardised journal impact factor") + 
  ylab("Coefficient of homophily (\u03B1')") + 
  theme_minimal()
ggsave(figure4, file = "figures/figure4.pdf", height = 5, width = 5, device = cairo_pdf)
saveRDS(R2, "manuscript/R2_for_figure4.RData")
figure4
```
<br></br>
**Figure `r figure.counter`**: Journal impact factor (expressed relative to the average for the discipline) is negatively correlated with $\alpha'$. The relationship is weak ($R^2$ = `r R2`), but the results suggest that journals with strong homophily tend to have lower impact factors than journals with weak homophily in the same discipline.

@@

### Gender homophily is also observed within countries

@@

```{r}
country.histo <- country.data %>%
  mutate(Significant = 1 * (p.adjusted < 0.05),
         Significant = replace(Significant, Significant == 0, "No"),
         Significant = replace(Significant, Significant == 1, "Yes")) %>%
  ggplot(aes(x = alpha.adjusted, fill = Significant)) + 
  geom_vline(xintercept = 0, linetype = 2) +
  geom_histogram(binwidth = 0.02, colour = "black") + 
  scale_fill_manual(values = c("skyblue", "white")) +
  xlab("Coefficient of homophily (\u03B1')") + 
  ylab("Number of journals") +
  theme_minimal() + 
  theme(legend.position = "top")
country.histo
```
<br></br>
**Figure S`r supp.figure.counter`**: Histogram of $\alpha'$ for 325 unique combinations of journal and country, using data from August 2015 - August 2016. The white areas denote combinations for which $\alpha'$ differs significantly from zero (p < 0.05, following false discovery rate correction).
